---
title: "Covid Data Analysis"
author: "pablo"
date: "4/29/2023"
output:
  pdf_document: default
  word_document: default
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Importing the Data

We read the data directly from the below URL, ad display the frist rows to get an idea of the schema. 

```{r }
library(tidyverse)
pth<-"https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/"

filenames<-c("time_series_covid19_confirmed_US.csv",  "time_series_covid19_confirmed_global.csv", "time_series_covid19_deaths_US.csv",  "time_series_covid19_deaths_global.csv")

us_cases<-read_csv(paste(pth,filenames[1],sep=''))
global_cases<-read_csv(paste(pth,filenames[2],sep=''))
us_deaths<-read_csv(paste(pth,filenames[3],sep=''))
global_deaths<-read_csv(paste(pth,filenames[4],sep=''))


global_cases <- global_cases %>%
  pivot_longer(cols=-c('Province/State','Country/Region','Lat','Long'),
               names_to='date',
               values_to='cases') %>%
  select(-c('Lat','Long'))

global_deaths <- global_deaths %>%
  pivot_longer(cols=-c('Province/State','Country/Region','Lat','Long'),
               names_to='date',
               values_to='deaths') %>%
  select(-c('Lat','Long'))

global <- global_cases %>%
  full_join(global_deaths) %>%
  rename(Country_Region=`Country/Region`,
         Province_State= `Province/State`) %>%
  mutate(date=mdy(date))



head(us_cases)
us_cases <- us_cases %>%
  pivot_longer(cols=-c(UID:Combined_Key),
               names_to='date',
               values_to='cases') %>%
  select(Admin2:cases) %>% 
  mutate(date=mdy(date)) %>%
  select(-c(Lat,Long_))
head(us_cases)

head(us_deaths)
us_deaths <- us_deaths %>%
  pivot_longer(cols=-c(UID:Combined_Key),
               names_to='date',
               values_to='deaths') %>%
  select(Admin2:deaths) %>% 
  mutate(date=mdy(date)) %>%
  select(-c(Lat,Long_))
head(us_deaths)

us<-us_cases %>%
  full_join(us_deaths)


global <- global %>%
  unite("Combined_Key",
        c(Province_State,Country_Region),
        sep=",",
        na.rm=TRUE,
        remove= FALSE)

uid<read_csv("https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/UID_ISO_FIPS_LookUp_Table.csv")
```
Next we check the class type of each column, 
```{r}
sapply(d,typeof)
```
We have to convert the dates from string to a date object
```{r}
d['OCCUR_DATE2'] <- as.Date(d$OCCUR_DATE,"%m/%d/%Y")
head(d$OCCUR_DATE2)
```

We see that most of our variables are categorical, so it is better to expore by using frequenices. 
We can ignore the coordinates as we won't conduct a geostatistical analysis.

```{r}
pct_na<-sapply(d,function(x){sum(is.na(x))/length(x)*100})
names <-names(d)

df<-data.frame(names,pct_na)
df
```
For the Categorical let's make a few frequencies

```{r,  echo=FALSE}
names(d[,sapply(d,function(x){typeof(x)=="character"})])

sapply(d[,c("BORO","LOC_OF_OCCUR_DESC","LOC_CLASSFCTN_DESC","LOCATION_DESC","STATISTICAL_MURDER_FLAG", "PERP_AGE_GROUP","PERP_SEX","PERP_RACE","VIC_AGE_GROUP","VIC_SEX","VIC_RACE")],function(x){t<-tibble::as.tibble(table(x)/27312);print(t) })


```

1.- We see that LOC_OF_OCCUR_DESC, LOC_CLASSFCTN_DESC have 93% missing so we can't use those columns. 
2.- Sex of the perpetrator is empty for 36%, but it is safe to impute M
3.- Sex of the victim has no missing values and 90% is male.

So we see an obvious pattern, that males are way overrepresented as victims ad pepetrators in this type of violent crime. Which matches our intuition



## Graphical Presentation of frequencies

Lets se the Frequencies Graphically 
```
library(ggplot2)
ggplot(data=data)+geom_bar(aes(x=BORO))+ggtitle("Borough")
ggplot(data=data)+geom_bar(aes(x=PERP_AGE_GROUP))+ggtitle("Perpetrator Age Group")

ggplot(data=data)+geom_bar(aes(x=PERP_RACE))+ggtitle("Perpetrator Race")+coord_flip()

ggplot(data=data)+geom_bar(aes(x=VIC_RACE))+ggtitle("Victim Race")+coord_flip()
```

```{r pressure, echo=FALSE}
d['year']<- sapply(d['OCCUR_DATE2'],year)
ggplot(data=d)+geom_bar(aes(x=year))+ggtitle("Year")
```


## Seasonality

Lets group the events by month and see if there are any patterns
```{r, echo=FALSE}
d['df']<-floor_date(d$OCCUR_DATE2,"month")
ts2<- d %>% 
  group_by(df) %>%
  summarise( n = n())
head(ts2)
ggplot(data=ts2,aes(x=df,y=n))+geom_line()
d['mth']<-month(floor_date(d$OCCUR_DATE2,"month"),label=TRUE)
ts2<- d %>% 
  group_by(mth ) %>%
  summarise( n = n()) 

head(ts2)
ggplot(data=ts2)+geom_bar(aes(x=mth,y=n),stat="identity")

```

And we see a spike in Summer.



We see a clear Seasonality Patter. 

## Test Seasonality

We will try to figure out the seasonality
```{r, echo=TRUE}

ts2<- d %>% 
  group_by(df ) %>%
  summarise( n = n()) 
head(ts2)

ts3<-ts(ts2$n,frequency = 12,start=c(2006,1))
plot(ts3)

ts_components <- decompose(ts3)
plot(ts_components)
summary(ts_components)
```

Notice in trend the the downward slope and the structural breakdown due to COVID which caused asharp increase and brought us back to pre 2010 levels. 
We see a strong seasonality component 

Below, we fit an arima model and we can see that the seasonal component is h
```{r }
###

library(forecast)
tsdisplay(ts3)

plot(acf(ts3))
plot(pacf(ts3))

model<-auto.arima(ts3,seasonal = TRUE)
plot(model$fitted)
lines(ts3, col='red')
qqnorm(model$residuals)
summary(model)

```
