---
title: "Covid Data Analysis"
author: "pablo"
date: "4/29/2023"
output:
  pdf_document: default
  word_document: default
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Importing the Data

We read the data directly from the below URL, ad display the frist rows to get an idea of the schema. 

```{r }
library(tidyverse)
pth<-"https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/"

filenames<-c("time_series_covid19_confirmed_US.csv",  "time_series_covid19_confirmed_global.csv", "time_series_covid19_deaths_US.csv",  "time_series_covid19_deaths_global.csv")

#Read Files from URL
us_cases<-read_csv(paste(pth,filenames[1],sep=''))
global_cases<-read_csv(paste(pth,filenames[2],sep=''))
us_deaths<-read_csv(paste(pth,filenames[3],sep=''))
global_deaths<-read_csv(paste(pth,filenames[4],sep=''))

# Transpose Global Cases
global_cases <- global_cases %>%
  pivot_longer(cols=-c('Province/State','Country/Region','Lat','Long'),
               names_to='date',
               values_to='cases') %>%
  select(-c('Lat','Long'))

#Transpose Global Deaths
global_deaths <- global_deaths %>%
  pivot_longer(cols=-c('Province/State','Country/Region','Lat','Long'),
               names_to='date',
               values_to='deaths') %>%
  select(-c('Lat','Long'))

#Merge Global Cases and Deaths
global <- global_cases %>%
  full_join(global_deaths) %>%
  rename(Country_Region=`Country/Region`,
         Province_State= `Province/State`) %>%
  mutate(date=mdy(date))



#Transpose US cases
us_cases <- us_cases %>%
  pivot_longer(cols=-c(UID:Combined_Key),
               names_to='date',
               values_to='cases') %>%
  select(Admin2:cases) %>% 
  mutate(date=mdy(date)) %>%
  select(-c(Lat,Long_))
head(us_cases)

#Transpose US deaths
head(us_deaths)
us_deaths <- us_deaths %>%
  pivot_longer(cols=-c(UID:Population),
               names_to='date',
               values_to='deaths') %>%
  select(Admin2:deaths) %>% 
  mutate(date=mdy(date)) %>%
  select(-c(Lat,Long_))
head(us_deaths)

#Merge US cases and deaths
us<-us_cases %>%
  full_join(us_deaths)
head(us)

#Create a combined key for global
global <- global %>%
  unite("Combined_Key",
        c(Province_State,Country_Region),
        sep=",",
        na.rm=TRUE,
        remove= FALSE)
#Import Population
uid<-read_csv("https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/UID_ISO_FIPS_LookUp_Table.csv") %>%
  select(-c(Lat, Long_,Combined_Key,code3,iso2,iso3,Admin2))
head(uid)

global<-global %>%
  left_join(uid,by=c('Province_State','Country_Region')) %>%
  select(-c(UID,FIPS)) %>%
  select(c(Province_State,Country_Region,date, cases,deaths,Population,Combined_Key))

head(global)
```


## Visualization 

```{r, echo=FALSE}
US_by_state <- us %>%
  group_by(Province_State,Country_Region,date) %>%
  summarize(cases =sum(cases), deaths=sum(deaths), Population=sum(Population)) %>%
  mutate(deaths_per_mill = deaths*1e6/Population) %>%
  select(Province_State, Country_Region, date, cases,deaths,deaths_per_mill,Population) %>%
  ungroup()

head(US_by_state)
```

```{r, echo=FALSE}
US_totals <- US_by_state %>%
  group_by(Country_Region,date) %>%
    summarize(cases =sum(cases), deaths=sum(deaths), Population=sum(Population)) %>%
  mutate(deaths_per_mill = deaths*1e6/Population) %>%
  select(Country_Region, date, cases,deaths,deaths_per_mill,Population) %>%
  ungroup()
head(US_totals)
```


Let's see the cumulative Covid Cases and Deaths in the US through time
```{r, echo=FALSE}
US_totals %>%
  filter(cases>0) %>%
  ggplot(aes(x=date,y=cases))+
  geom_line(aes(color="cases"))+
  geom_point(aes(color="cases"))+
  geom_line(aes(y=deaths, color="deaths"))+
  geom_point(aes(y=deaths, color="deaths"))+
  scale_y_log10()+
  theme(legend.position = "bottom",
        axis.text.x=element_text(angle=90)) +
  labs(title= "Covid19 in the US", y=NULL)

```



Now, for a specific state, lets say New Jersey

```{r, echo=FALSE}
state<- "New Jersey"
US_by_state %>%
  filter(Province_State==state)%>%
  filter(cases>0) %>%
  ggplot(aes(x=date,y=cases))+
  geom_line(aes(color="cases"))+
  geom_point(aes(color="cases"))+
  geom_line(aes(y=deaths, color="deaths"))+
  geom_point(aes(y=deaths, color="deaths"))+
  scale_y_log10()+
  theme(legend.position = "bottom",
        axis.text.x=element_text(angle=90)) +
  labs(title= str_c("Covid19 in ",state), y=NULL)

```


#Model: 

We are going to see if there is a significant difference of deaths per case across states in the US
To do so, lets fit an linear model for death as a function of cases with Province/State as a factor. 

``` {r, echo=FALSE}

US_by_state2 <- us %>%
  group_by(Province_State) %>%
  summarize(cases =sum(cases), deaths=sum(deaths)) %>%
  mutate(deaths_per_case = deaths/cases) %>%
  select(Province_State, deaths_per_case) %>%
    ungroup() %>%
  arrange(desc(deaths_per_case), .by_group = TRUE) 

f<-factor(US_by_state2$Province_State,levels=US_by_state2$Province_State)
US_by_state2$f <- f
ggplot(data=US_by_state2)+geom_bar(aes(x=f,y=deaths_per_case),stat="identity")+ coord_flip()
```

From the previous plot we expect Grand Princess and New Jersey to have the stronger effect. So these have the highest death per case. The coefficient in our Province_Region will adjust upward the relation

```{r, echo=FALSE}
library(mgcv)
model<- lm( deaths ~ cases + as.factor(Province_State), data=US_by_state)
summary(model)

coefficients <- model$coefficients

df<-data.frame(nm=names(coefficients), coefs=coefficients )
threshold<-mean(abs(coefficients))
ggplot(data=df[abs(coefficients)>threshold,])+geom_bar(aes(x=nm, y=coefs),stat="identity")+coord_flip()
```


After controlling for number of cases, the deadliest ones are NJ, NY and Texas.

# BIAS 

There are three sources of bias that i can identify, all related with the prevalence of testing.

1. The first is fundamental almost tautological, the results depend on the prevalence of testing. If you have low testing rates then you are most likely testing the patients that have worse prognosis, so the deaths per confirmed case will increase. 

2. In the US the testing/vaccines got politicized and response to the pandemic got split across political spectrum like red states vs blue states.

3. Wealthier, mega cities in both coasts have very high density which helps transmission and overwhelmed the medical response.